---
title: "ENV S 193DS"
subtitle: "Final"
format: pdf
author: Liv Zou
date: "2026-06-08"
---

Setup chunk:

```{r}
#| label: pkg
#| message: false
library(tidyverse)
library(dplyr)
library(janitor)
library(here)
install.packages("DHARMa")
library("DHARMa")
install.packages("ggeffects")
library(ggeffects)
```

# Problem 1 

1a. 

He most likely used a logistic regression because they were testing whether the distant to water edge predicted for the probability of microhabitat use. In the second part, he most likely perform a chi square test of independence because they were testing whether bird species and habitat (which are categorical variables) were related. 

1b. 

Part A: 

The American Avocets were more or less likely to use a microhabitat depending on their distance from the water. Distance to water's edge is a significant factor that influences microhabitat use. 

Part B: 

Use of habitat showed to be different across Avocets, Forster's Terns, and Black-necked Stilts. The three species showed unique patterns in the use of managed ponds, marshes, salt ponds, and mudflats. 

1c. 

An additional variable that could contribute to the probability of Avocet microhabitat use is human activity. Maybe it could be measured as a categorical variable organized in low, medium, and high levels, assessed by how many people are present in the habitat hourly. The birds may avoid those areas with more human activity because it might affect how much the birds go out of their shelters/nests to go eat, or may trigger them to fly out of their habitat when people are really close to them. 

# Problem 2

2a.

```{r}
#| message: false
nests <- read_csv(here("data", "nest_data_final.csv"))
```

```{r}
# Clean dataset
nests_clean <- nests |> 
  # Select relevant columns according to metadata 
     select(ant_species, height_cm, alive) |> 
  # Select the relevant ant species 
filter(ant_species %in% c("AB", "RRB", "BBR")) |> 
mutate(
    alive = case_when(
      alive == "A" ~ 1,
      alive == "D" ~ 0
    ),
  # Reorder ant species 
  # Set desired order
    ant_species = factor(ant_species,
                         levels = c("AB", "RRB", "BBR")),
    # Creating new columns displaying ant names 
    scientific_name = case_when(
      ant_species == "AB"  ~ "C. sjostedti",
      ant_species == "RRB" ~ "C. mimosae",
      ant_species == "BBR" ~ "C.nigriceps"))
```

```{r}
#| message: false
# Convert data into string and display structure
str(nests_clean)
```

```{r}
# Take 10 random samples
slice_sample(nests_clean, n = 10)
```

2b.

In the case column, a value of 1 means that a bird nest was present in the tree habitat while 0 indicates the absence of a bird nest. The values 1 and 0 represent if a tree was used as a nesting site or not. 

2c. 

The predictor variables are tree height and acacia ant species.

Tree height is a continuous quantitative variable, because it can be measured. As the height increases the probability of more nests appearing is expected to increase because the trees get bigger and taller, providing more spaces for birds to form shelteres that are not in the proximity of any predators that could probably harm them. This information is supported in the paper's results section where the probability of a bird deciding to nest increased when the trees increased by a 1 meter interval.  

Ant species is a categorical variable, because it has unique categories of something, in this case, there are distinctive, unique species of ants. Ant species may also influence the probability of nest appearance, and the abstract and introduction of Lujan's paper describes the relationship of ants and birds to be mutualistic. 

2d. 

```{r}
#| label: table showing only
#| echo: false
table <- tibble(
  model_number = c(1, 2),
  model_description = c(
    "Nest occurrence predicted by height and species without interactions",
    "Nest occurrence predicted by height species with interaction btwn variables."
  ),
  predictors = c(
    "height_cm + ant_species",
    "height_cm * ant_species"
  ),
  interaction_term = c("No", "Yes")
)

knitr::kable(table)
```

2e. 

```{r}
#| message: false
table <- tibble(
  model_number = c(1, 2),
  model_description = c(
    "Nest occurrence predicted by height and species without interactions",
"Nest occurrence predicted by height and species w/ interaction btwn variables."
  ),
  predictors = c(
    "height_cm + ant_species",
    "height_cm * ant_species"
  ),
  interaction_term = c("No", "Yes")
)
```

```{r}
model1 <- glm(
 alive ~ height_cm + ant_species,
  data = nests_clean,
  family = binomial
)
```

```{r}
model2 <- glm(
alive ~ height_cm * ant_species,
  data = nests_clean,
  family = binomial
)
```

2f.

```{r}
AIC(model1, model2)
```

The best model as determined by Akaike's Information Criterion (AIC) is one that includes tree height and ant species as predictor variables. The model has a lower AIC (20.52) compared to the model that included the interaction between height and ant species (24.52) which had the interaction between tree height and ant species. This implies that the model selected is better in explaining and predicting the data as well without including unnecessary complexity in the visualization.

2g. 

```{r}
model1_residual <- simulateResiduals(fittedModel = model1)
plot(model1_residual)
```

2h. 


```{r}
predictions1 <- ggpredict(model1, terms = c("height_cm [all]", "ant_species"))
predictions1 <- as.data.frame(predictions1)

ggplot() + 
  geom_point(data = nests_clean, aes(x = height_cm, y = alive), 
             color = "darkblue", alpha = 0.5, size = 2) + 
  geom_ribbon(data = predictions1, aes(x = x, ymin = conf.low, ymax = conf.high), fill = "lightgreen", 
              alpha = 0.2) + 
  geom_line(data = predictions1, aes(x = x, y = predicted), color = "red", linewidth = 1) +
  facet_wrap(~ group) + labs(x = "Nest height above ground (cm)", y = "Probability of survival", title = "Model predictions of ant survival determined by nest distance and ant species")
```

2i.

Figure 1. Model predictions of ant survival across species and nest height. Probabilities of survival were estimated using AIC and linear regression that included nest height above the ground, and various species of ants. The dark blue points represent the amount of ant that survived per species while the ones on the bottom are representing the ones that did not survive. The line represents the model predictions of ant survival. The green shaded regions represent the 95% confidence interval. The results are distributed across each ant species. This information was extracted from the nests_clean dataset. 

2j. 

```{r}
predict2j <- predict(
  model1,
  newdata = data.frame(
    height_cm = c(600, 600, 600),
    ant_species = factor(
      c("RRB", "AB", "BBR"),
      levels = levels(nests_clean$ant_species)
    )
  ),
  type = "response"
)
predict2j
```


2k. 

I noticed that nest occupancy was higher in taller trees. At 600 cm, the model predicted high likelihood of nest occurrences across the three species of ants, the probability ranging from 0.984 for Mimosae ants to 1 for both Sjostedti and Nigriceps ants. These predictions suggest that height and types of ants are influential variables in nest appearance, and that taller trees have more nest occupancy. This makes sense because taller trees are higher off the ground, away from many predators, and provide more shade as there are more leaves as the nest is higher, providing safety, shelter and shade for the birds that create their nests. Additionally, higher trees are also farther away from any human disturbances. Ant species being an influential factor for nest occupancy likelihood is also relevant. Though the relationship between ants and nesting may be mutualistic, each ant may have different ways of defending nests when they are on the trees, making some habitats more favorable to birds than others. 

# Problem 3

3a. 

My visualizations in Homework 2 were very different because they 


